---
title: "Practical Machine Learning Course Project"
author: "P. Lombardo"
date: "7/16/2020"
output: html_document
---

```{r ref.label = package_load, echo=FALSE, results='hide', message = FALSE}
library(dplyr);library(ggplot2);library(caret)
```






### Resources

[http://pjl414.github.io/practicalmachinelearning](https://pjl414.github.io/practicalMachineLearning/)

[guide](https://github.com/lgreski/datasciencectacontent/blob/master/markdown/pml-ghPagesSetup.md)

```{r}
urltrain<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urltest<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download.file(urltrain,
#               destfile = "data/train.csv",
#               method = "curl")
# download.file(urltest,
#               destfile = "data/test.csv",
#               method = "curl")
str(train)
```

```{r}
train<-read.csv("data/train.csv", header = T)[,-1]
length(names(train))

collect_numerics<-character()
collect_chars<-character()
for (i in 1:length(names(train))){
    nm<-names(train)[i]
    if (is.numeric(train[,nm])){
        collect_numerics<-c(collect_numerics,nm)
    } else {
        collect_chars<-c(collect_chars,nm)
    }
}
collect_numerics
train_num<-train %>% select(all_of(collect_numerics),classe)
train_char<-train %>% select(all_of(collect_chars))
```
158 different possible predictors; we're attempting to classify into 5 different classes:
```{r}
train$classe %>% unique()
train$classe %>% class()
train$classe<-as.factor(train$classe)
```
Probably shouldn't use `user_name` as a predictor since we want to predict on new users?

Many NA values, perhaps do some data imputation to get values for these?
    * perhaps grouping by some of the naming conventions (i.e. where sensors were place) would be a good idea for helping with data imputation.

```{r}
train$classe %>% unique()
```
```{r}
names(train)
```
Group namings?

* 



## Citations:
Many thanks to the researchers of the following work for sharing their data:

> Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6.

> Read more: [http://groupware.les.inf.puc-rio.br/har#sbia_paper_section#ixzz6SOLLlRhI](http://groupware.les.inf.puc-rio.br/har#sbia_paper_section#ixzz6SOLLlRhI)


